Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-04-26 manuscript/manuscript.tex
final proofread of the article

\documentclass[twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\usepackage{lmodern} % Chloe: better font rendering 
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Yet, network methods are notably unstable, producing different results when the input data changes. Hence, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P-value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, a ubiquitin ligase related that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. based on logistic regression). Those SNPs with a P-value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

@@

In this work, we analyze the effectiveness of these six network methods for biomarker discovery on GWAS data. While all of them capture susceptibility mechanisms resembling that postulated by the omnigenic model, they do so in different ways, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network-based methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy--Weinberg equilibrium in controls (P-value \textless~0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After controlling for relatedness, 17 additional samples were removed (6 for sample identity error, 6 controls related to other samples, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with lobular carcinoma \emph{in situ}, 1 with a \emph{BRCA1/2} mutation detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

@@

\subsection{High-score subnetwork search algorithms}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1~d.f.~\(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 \cite{mishra_vegas2:_2015} to compute the gene-level association score from the P-values of the SNPs mapped to them. Specifically, for each gene we only used the 10\% of SNPs mapped to it with lowest P-values. We mapped SNPs to genes through their genomic coordinates: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the 197\,083 SNPs remaining after quality control, 164\,037 were mapped to at least one of these genes.

@@

In addition, we use several matrices that describe different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_i$ and $v_j$. The \emph{adjacency matrix} $W_G$ contains a 1 when the corresponding nodes are connected, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_G$ is a diagonal matrix which contains the degree of the different nodes.

@@

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only network method designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this work better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association scores of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}). The properties of these three subnetworks is available in Supplementary Table \ref{tab:snp_solutions}.

@@

\begin{table}[htbp]
  \caption{Summary of the differences between the network methods.}
  \label{tab:method_comparison}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Comp. & Input & Scoring & Ref.\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & 1 d.f. \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & \(\ge\) 1 & Summary & $\Phi$\textsuperscript{-1}(1 - P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \footnotesize{
        \item \textbf{\emph{Field}}: field in which the algorithm was developed. \textbf{\emph{Nodes}}: the type of network, either gene (protein-protein interaction network usually) or a SNP network. \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored. \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected. \textbf{\emph{Comp.}}: number of connected components in the solution. \textbf{\emph{Input}}: genotype data or GWAS summary statistics. \textbf{\emph{Scoring}}: how SNP/gene P-values are transformed into node scores. In the case of SigMod, $\Phi$\textsuperscript{-1} represents the inverse of the
        cumulative distribution function of the standard Normal distribution. \textbf{\emph{Ref.}}: original publication featuring the algorithm.
      }
    \end{tablenotes}
  \end{threeparttable}
\end{table}

Genes that contribute to the same function are nearby in the PPIN, might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). But this is not the only aspect to model when developing a network method: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, which network to use, etc. Unsurprisingly, multiple solutions have been proposed. We examined six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table~\ref{tab:method_comparison}.

\begin{description}
\item[{dmGWAS}] dmGWAS seeks the subgraph with the highest local density in low P-values \cite{jia_dmgwas:_2011}. To that end it searches candidate subnetwork solutions using a greedy, ``seed and extend'', heuristic:

\begin{enumerate}
\item Select a seed node $i$ and form the subnetwork $S_i=\{i\}$.
\item Compute Stouffer's Z-score Z\textsubscript{m} for the current subgraph $S_i$ as

\begin{equation*} 
Z_m = \frac{1}{\sqrt{k}} \sum_{j \in S_i} z_j
\end{equation*}

where \emph{k} is the number of genes in $S_i$; z\textsubscript{j} is the Z score of gene $j$, computed as \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{j}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes of $S_i$,  i.e. nodes at distance \(\le\) d.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than a threshold Z\textsubscript{m}~\texttimes{}~(1 + r).
\item Repeat 2-4 until no further enlargement is possible.
\item Add $S_i$ to the list of subnetworks to return. Its Z-score is normalized as

\begin{equation*}
Z_{N}=\frac{Z_{m}-\operatorname{mean}\left(Z_{m}(\pi)\right)}{\operatorname{SD}\left(Z_{m}(\pi)\right)}
\end{equation*}

@@

DmGWAS carries out this process on every gene in the network. We used the implementation of dmGWAS in the dmGWAS 3.0 R package \cite{dmgwas}. We used the suggested hyperparameters $d = 2$ and $r = 0.1$. We used the function \emph{simpleChoose} to select the solution subnetwork, which aggregates the top 1\% modules into the solution subnetwork.
\end{description}

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P-value into a score that is negative under no association with the phenotype, and positive when there is. This transformation is achieved by modelling the distribution of P-values by a beta-uniform model (BUM) parameterized by the desired false discovery rate (FDR). Thus formulated, the problem is NP-complete, and hence solving it would require a prohibitively long computational time. To solve it efficiently it is re-cast as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S), defined as:

@@

\item[{HotNet2}] HotNet2 was developed to find connected subgraphs of genes frequently mutated in cancer \cite{leiserson_pan-cancer_2015}. To that end, it considers both the local topology of the network and the scores of the nodes. The former is captured by an insulated heat diffusion process: at initialization, the score of the node determines its initial heat; iteratively each node yields heat to its ``colder'' neighbors, and receives heat from its ``hotter'' neighbors, while retaining part of its own (hence, \emph{insulated}). This process continues until equilibrium is reached, and results in a diffusion matrix $F$. $F$ is used to compute the similarity matrix $E$ that models exchanged heat as
\begin{equation*} 
E = F \operatorname{diag}(w(V)), 
\end{equation*}
where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. For any two nodes/genes $i$ and $j$, $E_{ij}$ models the amount of heat that diffuses from node $j$ to node $i$, which can be interpreted as a (non-symmetric) similarity between those two nodes. To obtain densely connected subnetworks, HotNet2 prunes $E$, only preserving edges such that $w(E) > \delta$. Lastly, HotNet2 evaluates the statistical significance of the subnetworks by comparing their size to the size of networks obtained by permuting the node scores. We assigned initial node scores as in \citet{nakka_gene_2016}, assigning a score of 0 for the genes with low probability of being associated to the disease, and -log\textsubscript{10}(P-value) to those likely to be. In this dataset, the threshold separating both was a P-value of 0.125, which was obtained using a local FDR approach \cite{scheid_twilight;_2005}. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, which is robust to their values \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

\item[{LEAN}] LEAN searches altered ``star'' subnetworks, that is, subnetworks composed by one central node and all its interactors \cite{gwinner_network-based_2016}. By imposing this restriction, LEAN is able to exhaustively test all such subnetworks (one per node). For a particular subnetwork of size \emph{m}, the P-values corresponding to the involved nodes are ranked as p\textsubscript{1} \(\le\) \dots{} \(\le\) p\textsubscript{m}. Then, \emph{k} binomial tests are conducted, to compute the probability of having \emph{k} out of \emph{m} P-values lower or equal to p\textsubscript{k} under the null hypothesis. The minimum of these \emph{k} P-values is the score of the subnetwork. This score is transformed into a P-value through an empirical distribution obtained via a subsampling scheme, where gene sets of the same size are selected randomly, and their score computed. Lastly, P-values are corrected for multiple testing through a Benjamini-Hochberg correction. We used the implementation of LEAN from the LEANR R package \cite{leanr}.
\item[{SConES}] SConES searches the minimal, modular, and maximally associated subnetwork in a SNP graph \cite{azencott_efficient_2013}. Specifically, it solves the problem

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. The connectivity term penalizes disconnected solutions, with many edges between a node that is selected and a node that is not. Given a $\lambda$ and an $\eta$, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. In this case, the selected hyperparameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI. We used the version on SConES implemented in the R package martini \cite{martini}.
\end{description}

@@

\begin{description}
\item[{Consensus}] In addition, we built a consensus network by retaining the nodes that were selected by at least two of the six methods (using SConES GI for SConES).
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P-value \textless~0.05 were deemed significant. As the significant pathways are often overlapping and redundant, the results were manually curated afterwards.

@@

We evaluated multiple properties of the different method through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the standard deviation of the different measures.

@@

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input, here, using different subsets of the data) of the different network methods. The stability was quantified using the Pearson correlation to measure the overlap between different runs as suggested by \citet{nogueira_measuring_2016}.

@@

In terms of computational runtime, the fastest method was heinz (Figure~\ref{fig:benchmark}D), which finds the solution efficiently in a few seconds. HotNet2 was the slowest (3 days and 14 hours on average). Including the time required to compute the gene scores, however, slows down considerably gene-based methods; on this benchmark, that step took on average 1 day and 9.33 hours. Considering that, it took 5 days on average for HotNet2 to produce a result.

>>>>

\documentclass[twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\usepackage{lmodern} % Chloe: better font rendering
\usepackage[shortcuts]{extdash} 
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women with a family history. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable, producing different results when the input data changes slightly. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, a ubiquitin ligase related that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. based on logistic regression). Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

@@

In this work, we analyze the effectiveness of these six network methods for biomarker discovery on GWAS data. While all of them capture susceptibility mechanisms resembling that postulated by the omnigenic model, they do so in different ways, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy--Weinberg equilibrium in controls (P\=/value \textless~0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After controlling for relatedness, 17 additional samples were removed (6 for sample identity error, 6 controls related to other samples, 2 cases being related to an index case, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with lobular carcinoma \emph{in situ}, 1 with a \emph{BRCA1/2} mutation detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

@@

\subsection{High-score subnetwork search algorithms}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1~d.f.~\(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2 \cite{mishra_vegas2:_2015} to compute the gene-level association score from the P\=/values of the SNPs mapped to them. Specifically, for each gene we only used the 10\% of SNPs mapped to it with lowest P\=/values. We mapped SNPs to genes through their genomic coordinates: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the 197\,083 SNPs remaining after quality control, 164\,037 were mapped to at least one of these genes.

@@

In addition, we use two matrices that describe different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_i$ and $v_j$. The \emph{adjacency matrix} $W_G$ contains a 1 when the corresponding nodes are connected, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_G$ is a diagonal matrix which contains the degree of the different nodes.

@@

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only network method designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this work better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association scores of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}). The properties of these three subnetworks are available in Supplementary Table \ref{tab:snp_solutions}.

@@

\begin{table}[htbp]
  \caption{Summary of the differences between the network methods.}
  \label{tab:method_comparison}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Comp. & Input & Scoring & Ref.\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & 1 d.f. \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & \(\ge\) 1 & Summary & $\Phi$\textsuperscript{-1}(1 - P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \footnotesize{
        \item \textbf{\emph{Field}}: field in which the algorithm was developed. \textbf{\emph{Nodes}}: the type of nodes in the network, either genes (PPIN) or SNPs. \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored. \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than containing high scoring, connected nodes. \textbf{\emph{Comp.}}: number of connected components in the solution. \textbf{\emph{Input}}: genotype data or GWAS summary statistics. \textbf{\emph{Scoring}}: how SNP/gene P\=/values are transformed into node scores. In the case of SigMod, $\Phi$\textsuperscript{-1} represents the inverse of the
        cumulative distribution function of the standard Normal distribution. \textbf{\emph{Ref.}}: original publication featuring the algorithm.
      }
    \end{tablenotes}
  \end{threeparttable}
\end{table}

Genes that contribute to the same function are nearby in the PPIN, and can be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). But this is not the only aspect to model when developing a network method: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, which network to use, etc. Unsurprisingly, multiple solutions have been proposed. We examined six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table~\ref{tab:method_comparison}.

\begin{description}
\item[{dmGWAS}] dmGWAS seeks the subgraph with the highest local density in low P\=/values \cite{jia_dmgwas:_2011}. To that end it searches candidate subnetwork solutions using a greedy, ``seed and extend'', heuristic:

\begin{enumerate}
\item Select a seed node $i$ and form the subnetwork $S_i=\{i\}$.
\item Compute Stouffer's Z-score Z\textsubscript{m} for $S_i$ as

\begin{equation*} 
Z_m = \frac{1}{\sqrt{k}} \sum_{j \in S_i} z_j,
\end{equation*}

where \emph{k} is the number of genes in $S_i$; z\textsubscript{j} is the Z score of gene $j$, computed as \(\phi\)\textsuperscript{-1}(1 - P\=/value\textsubscript{j}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes of $S_i$,  i.e. nodes at distance \(\le\) d.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than a threshold Z\textsubscript{m}~\texttimes{}~(1 + r).
\item Repeat 2-4 until no further enlargement is possible.
\item Add $S_i$ to the list of subnetworks to return. Its Z-score is normalized as

\begin{equation*}
Z_{N}=\frac{Z_{m}-\operatorname{mean}\left(Z_{m}(\pi)\right)}{\operatorname{SD}\left(Z_{m}(\pi)\right)},
\end{equation*}

@@

DmGWAS carries out this process on every gene in the network. We used the implementation of dmGWAS in the dmGWAS 3.0 R package \cite{dmgwas}. We used the suggested hyperparameters $d = 2$ and $r = 0.1$. We used the function \emph{simpleChoose} to select the solution subnetwork, which aggregates the top 1\% subnetworks.
\end{description}

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P\=/value into a score that is negative under no association with the phenotype, and positive when there is. This transformation is achieved by modelling the distribution of P\=/values by a beta-uniform model (BUM) parameterized by the desired false discovery rate (FDR). Thus formulated, the problem is NP-complete, and hence solving it would require a prohibitively long computational time. To solve it efficiently it is re-cast as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S), defined as:

@@

\item[{HotNet2}] HotNet2 was developed to find connected subgraphs of genes frequently mutated in cancer \cite{leiserson_pan-cancer_2015}. To that end, it considers both the local topology of the network and the scores of the nodes. The former is captured by an insulated heat diffusion process: at initialization, the score of the node determines its initial heat; iteratively each node yields heat to its ``colder'' neighbors, and receives heat from its ``hotter'' neighbors, while retaining part of its own (hence, \emph{insulated}). This process continues until equilibrium is reached, and results in a diffusion matrix $F$. $F$ is used to compute the similarity matrix $E$ that models exchanged heat as
\begin{equation*} 
E = F \operatorname{diag}(w(V)), 
\end{equation*}
where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. For any two nodes $i$ and $j$, $E_{ij}$ models the amount of heat that diffuses from node $j$ to node $i$, which can be interpreted as a (non-symmetric) similarity between those two nodes. To obtain densely connected subnetworks, HotNet2 prunes $E$, only preserving edges such that $w(E) > \delta$. Lastly, HotNet2 evaluates the statistical significance of the subnetworks by comparing their size to the size of networks obtained by permuting the node scores. We assigned initial node scores as in \citet{nakka_gene_2016}, assigning a score of 0 for the genes with low probability of being associated to the disease, and -log\textsubscript{10}(P\=/value) to those likely to be. In this dataset, the threshold separating both was a P\=/value of 0.125, which was obtained using a local FDR approach \cite{scheid_twilight;_2005}. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, which is robust to their values \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

\item[{LEAN}] LEAN searches altered ``star'' subnetworks, that is, subnetworks composed by one central node and all its interactors \cite{gwinner_network-based_2016}. By imposing this restriction, LEAN is able to exhaustively test all such subnetworks (one per node). For a particular subnetwork of size \emph{m}, the P\=/values corresponding to the involved nodes are ranked as p\textsubscript{1} \(\le\) \dots{} \(\le\) p\textsubscript{m}. Then, \emph{k} binomial tests are conducted, to compute the probability of having \emph{k} out of \emph{m} P\=/values lower or equal to p\textsubscript{k} under the null hypothesis. The minimum of these \emph{k} P\=/values is the score of the subnetwork. This score is transformed into a P\=/value through an empirical distribution obtained via a subsampling scheme, where gene sets of the same size are selected randomly, and their score computed. Lastly, P\=/values are corrected for multiple testing through a Benjamini-Hochberg correction. We used the implementation of LEAN from the LEANR R package \cite{leanr}.
\item[{SConES}] SConES searches the minimal, modular, and maximally associated subnetwork in a SNP graph \cite{azencott_efficient_2013}. Specifically, it solves the problem

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. The connectivity term penalizes disconnected solutions, with many edges between nodes that are selected and nodes that are not. Given a $\lambda$ and an $\eta$, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. In this case, the selected hyperparameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI. We used the version on SConES implemented in the R package martini \cite{martini}.
\end{description}

@@

\begin{description}
\item[{Consensus}] We built a consensus network by retaining the nodes that were selected by at least two of the six methods (using SConES GI for SConES).
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P\=/value \textless~0.05 were deemed significant. As the significant pathways are often overlapping and redundant, the results were manually curated afterwards.

@@

We evaluated multiple properties of the different methods (described below) through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the standard deviation of the different measures.

@@

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input, here, using different \emph{train} sets) of the different network methods. The stability was quantified using the Pearson correlation to measure the overlap between different runs as suggested by \citet{nogueira_measuring_2016}.

@@

In terms of computational runtime, the fastest method was heinz (Figure~\ref{fig:benchmark}D), which returned a solution in a few seconds. HotNet2 was the slowest (3 days and 14 hours on average). Including the time required to compute the gene scores, however, slows down considerably gene-based methods; on this benchmark, that step took on average 1 day and 9.33 hours. Considering that, it took 5 days on average for HotNet2 to produce a result.
